Linting Linearity in Core/System FC

### 2023-01-11 report.tex
introduction: linearity in core

@@

As an example, consider the following linearly typed program which allocates
some memory and frees it twice. The variable $p$ is a linear resource created by
the \textbf{malloc} keyword.

\begin{code}
let p = malloc 4
 in free p
    free p
\end{code}

@@

\begin{itemize}
    \item Linear types were only added to the language roughly \emph{31 years
        after} Haskell's inception. It is an especially difficult endeavour to
        add linear types to a well-established language with a large library
        ecosystem and many active users, rather than to develop the language
        from the ground up with linear types in mind, and the successful
        approach as implemented in GHC 9.0, the leading Haskell compiler, was
        based on Linear Haskell\cite{linearhaskell}, where a linear type system
        designed with retaining backwards-compatibility and allowing code reuse
        across linear and non-linear users of the same libraries in mind was
        described. We describe Linear Haskell in detail in
        section~\ref{linear-haskell-section}.
    \item Linear types permeate Haskell down to (its) \textbf{Core}, the intermediate
        language Haskell is desugared to. \textbf{Core} is a very small
        explicitly typed language to which multiple Core-to-Core optimizing
        transformations are applied in succession. The final Core program can be
        typechecked very fast (because of the explicit type annotations) and
        doing so serves as a sanity check to the correction of the
        source transformations. If the resulting optimized Core program doesn't
        typecheck, something \emph{very wrong} has happened in a transformation!
        We discuss Core in detail in section~\ref{core-section}.
\end{itemize}

In theory, the linearity annotations in \textbf{Core} should allow for the same
typechecking verification we had with the existing type annotations prior to the
addition of linear types, that is, our Core program with linearity annotations
should be typechecked after the optimizing transformations...

>>>>

% Linear types are cool
% Among the few, Haskell has linear types
% But Haskell differs from other languages with linear types
%   * added 31 years after inception
%   * added to its core
% Core has linear types for good reasons
% But they are broken -- why?
%   * Optimizations changes source -- e.g. if recursive lets are introduced, how to type linearity?
%

@@

As an example, consider the following linearly typed program which allocates
some memory and frees it twice. The variable $p$ is a linear resource created by
the \textbf{malloc} keyword.
\begin{code}
let p = malloc 4
 in free p
    free p
\end{code}

@@

\begin{itemize}
    \item Linear types were only added to the language roughly \emph{31 years
        after} Haskell's inception. It is an especially difficult endeavour to
        add linear types to a well-established language with a large library
        ecosystem and many active users, rather than to develop the language
        from the ground up with linear types in mind, and the successful
        approach as implemented in GHC 9.0, the leading Haskell compiler, was
        based on Linear Haskell\cite{linearhaskell}, where a linear type system
        designed with retaining backwards-compatibility and allowing code reuse
        across linear and non-linear users of the same libraries in mind was
        described. We describe Linear Haskell in detail in
        section~\ref{linear-haskell-section}.
    \item Linear types permeate Haskell down to (its) \textbf{Core}, the intermediate
        language Haskell is desugared to. \textbf{Core} is a very small
        explicitly typed language to which multiple Core-to-Core optimizing
        transformations are applied in succession. The final Core program can be
        typechecked very fast (because of the explicit type annotations) and
        doing so serves as a sanity check to the correction of the
        source transformations. If the resulting optimized Core program doesn't
        typecheck, something \emph{very wrong} has happened in a transformation!
        We discuss Core in detail in section~\ref{core-section}.
    % TODO: \item values in rust are linear by default while non-linear is the
        % haskell default?
\end{itemize}

Core was imbued with linear types because an implementation detail required it,
and, more importantly, to guarantee linear programs remain linear after being
desugared and optimized: much the same way a \emph{typed} Core ensures that
desugaring, and optimizing transformations, are correctly implemented and result
in a \emph{well-typed} program -- a design decision that has well served the
compiler developers with a safety net for a long time. It is crucial that a
program behaviour enforced by linear types, regarding the usage of certain
resources, is \emph{not} changed in the desugaring or optimization process
(optimizations should not destroy linearity!) and a linearity aware Core
typechecker can guarantee that.

Additionally, while not yet a reality, linearity in Core could be used to inform
certain program optimizations, i.e. having linear types in Core could be used to
further optimize certain programs and, therefore, benefit the runtime
performance characteristics of our programs. For example, Linear Haskell\cite{}
describes as future work an improvement to the inlining optimization: Inlining
is a centerpiece program optimization primarily because of the subsequent
optimizing opportunities unlocked by inlining. However, it relies on a heuristic
process known as \emph{cardinality analysis} to discover safe inlining
opportunities. Unfortunately, heuristics can be volatile and fail in identifying
crucial inlining opportunities resulting in programs that fall short of their
performance potential. On the contrary, the linearity information could be
integrated in the analysis and used as a (non-heuristic) cardinality
\emph{declaration}.

However, currently, typechecking linearity in Core is broken in the face of
optimizations! The reason is not evidently clear: if we can typecheck linearity
in the surface level Haskell why can't we in Core? In fact, if we were to
typecheck the programs in Core as they were at the surface level, we would
succeed, but programs are \emph{transformed} (by semantic-preserving optimizing
transformations) and hence end up different from how they were originally
written. For example, the following recursive let definition, where $x$ is a
linear variable that must be used exactly once, would not typecheck in a source
linear Haskell program since the typechecker cannot tell that $x$ is used
linearly, but this program might occur naturally in Core from
transformations on a program that did succesfully typecheck:
\begin{code}
letrec f = \case
        True -> f False
        False -> x
in f True
\end{code}

Is this program really still linear? If one looks at it

% Alternatively, one might imagine Haskell being desugared into an intermediate
% representation to which multiple different optimizing transformations are
% applied but on which no integrity checks are done

Despite \emph{want} a linearly-typed core
because a linearly-typed Core ensures that desugaring
Haskell and optimizing transformations don't destroy linearity.

% In theory, because the Core language must also know about linearity, we should

% A remedy is to use the multiplicity annotations of λq → as cardinality declarations. Formalising
% and implementing the integration of multiplicities in the cardinality analysis is left as future work.

% In theory, we should typecheck \emph{linearity} in \textbf{Core} just the same
% as the typechecking verification we had with the existing type annotations prior
% to the addition of linear types, that is, our Core program with linearity
% annotations should be typechecked after the optimizing transformations...
